---
title: "From IDS to analysis"
author: "Göran Broström"
date: "May 7, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(knitr)
library(eha)
```

## Introduction

We have two data files with data in *IDS format*, *INDIVIDUAL.rda* and *INDIV_INDIV.rda*. From these files our task is to create a data file that is suitable for *statistical analysis*.

The result will be a file based on *children*, in a wide sense (all grown-up have been children earlier in life). You could also say that it will be based on *mothers*.

This is what we will accomplish: Family trees on the female side, built up from the roots consisting of all women without link to a biological mother. These "start atoms" we call *generation 0*. Next, *generation 1* consists of all offspring to generation 0, including males. Then, generation 2 consists of all offspring to the *females in generation 1*, also males, and so on. A branch is thus always finished when a male pops up. 

## The *INDIVIDUAL* data frame

Let us have a look at *INDIVIDUAL*:

```{r lookINDIVIDUAL}
load("data/INDIVIDUAL.rda")
kable(head(INDIVIDUAL), row.names = FALSE)
```

This is the total information on *id = 1*. The structure of the data frame is repeated for each individual: Six rows per individual, and only *static data* (fixed over time). This is a typical example of an *un-tidy data set*. A *tidy* data set has variables in columns and observations in rows, and that is what is needed in any structured statistical data analysis (source: [Hadley Wickham](https://www.jstatsoft.org/article/view/v059i10)). 

This un-tidiness is easily fixed, but we prefer to start in the other end, with the *INDIV_INDIV* data frame.

## The *INDIV_INDIV* data frame

The *INDIV_INDIV* data frame has a very simple structure:

```{r lookINDIV_INDIV}
library(xtable)
load("data/INDIV_INDIV.rda")
kable(head(INDIV_INDIV), row.names = FALSE)
```

The variable *id_1* is the mother of *id_2*, and the variable *Relation* is *redundant* (constant equal to "Mother"). *INDIV_INDIV* is a tidy data frame, if we regard *id_2* as the key of the *observations*.

## Creation of the analysis file

Remember, *id_2* is the child of *id_1*, so a starting point for our efforts is the data frame *per*, given by 

```{r getper}
per <- data.frame(id = INDIV_INDIV$id_2,
                  mid = INDIV_INDIV$id_1)
summary(per)
```
However, the individuals in *INDIVIDUAL* that have no mother (in INDIVIDUAL) must be added. Or, strictly, only the *females* without a mother. Or, even stricter, only those women who are mothers to persons in *INDIVIDUAL*. So,

```{r motherless}
atoms <- unique(INDIVIDUAL$id_I[!(INDIVIDUAL$id_I %in% per$id) & INDIVIDUAL$id_I %in% per$mid])
nat <- length(atoms)
```
So, there are `r nat` *atom mothers*. These women will constitute the roots of the family trees we will build. For now, we add them to *per*:

```{r addatom}
per <- rbind(per, data.frame(id = atoms, mid = rep(NA, nat)))
```

Next we pick the six static variables *sex*, *birthdate*, *deathdate*, *startdate*, *enddate*, and *marriagedate* from *INDIVIDUAL*:

```{r pickfrom}
Types <- unique(INDIVIDUAL$Type)
for (type in Types){
    indx <- match(per$id, INDIVIDUAL[INDIVIDUAL$Type == type, "id_I"])
    per[, type] <- INDIVIDUAL[INDIVIDUAL$Type == type, ]$Value[indx]
}

kable(head(per), row.names = FALSE)
```

Let's look at the characteristics of the data frame *per*:

```{r strper}
str(per)
```
As a consequence of the fact that *INDIVIDUAL* is an un-tidy data frame, the date and sex variables had to be given as *text strings*. We want *Sex* to be a *factor* and the dates are for our coming purposes best treated as numeric (unit: years after "00-01-01 BC", which does not exist, really). So

```{r convertvar}
per$Birth_date <- as.double(toTime(per$Birth_date))
per$Death_date <- as.double(toTime(per$Death_date))
per$Sex <- as.factor(per$Sex)
per$Start_observation <- as.double(toTime(per$Start_observation))
per$End_observation <- as.double(toTime(per$End_observation))
per$Marriage_date <- as.double(toTime(per$Marriage_date))
kable(head(per), row.names = FALSE)
```

Now we want the ages at which events occur, so

```{r getages}
per$enter <- per$Start_observation - per$Birth_date
per$exit <- per$End_observation - per$Birth_date
per$death <- per$Death_date - per$Birth_date
per$marriage <- per$Marriage_date - per$Birth_date
kable(head(per[, c("id", "mid", "Birth_date", "enter", "marriage", "exit", "death")]))
```

We further define the variable *event* as an indicator of *death before age one*:

```{r defeve}
per$event <- !is.na(per$death) & per$death <= 1
table(per$event)
```

## Generations

*Generation 0* consists of all mothers without a mother(!).

```{r gen0}
per$generation <- rep(NA, NROW(per))
per$generation[is.na(per$mid)] <- 0
with(per, table(generation, Sex, useNA = "ifany"))
```

*Generation 1* consists of all individuals whose mother is in Generation 0:

```{r}
gen0 <- per$id[!is.na(per$generation) & per$generation == 0]
per$generation[!is.na(per$mid) & per$mid %in% gen0] <- 1
with(per, table(generation, Sex, useNA = "ifany"))
```

We got the idea; the rest is executed in a *while* loop:

```{r exeforloop}
genX <- per$id[!is.na(per$generation) & per$generation == 1]
loo <- 1
gen <- 1
while (loo){
    oo <- !is.na(per$mid) & (per$mid %in% genX)
    loo <- sum(oo)
    if (loo){
        gen <- gen + 1
        per$generation[oo] <- gen
        genX <- per$id[!is.na(per$generation) & per$generation == gen]
    }
}
table(per$generation, per$Sex, useNA = "ifany")
```
So, there are 10014 "atom mothers", each the root of a family tree. The largest tree covers ten generations. Note that new branches of a tree is only generated by females; a male node denotes a dead end.

Now we need to put on *grandmother's id* on everyone in the second generation and above. Easy in principle: we have the mother's id of all in these generations and in generation 1, and grandmother is just mother's mother ...

```{r getgrandma}
per$gmid <- per$mid[match(per$mid, per$id)]
```

Have a look:

```{r havealook}
(x <- per[1, c("id", "mid", "gmid")])
per[per$id == x$mid, c("id", "mid", "gmid")]
```

We trust it worked (but the sample is small ...). Finally, we would like to put on a variable *tree*, equal to the id of the root, on each individual.

```{r putontree}
per$tree <- rep(NA, NROW(per))
per$tree[per$generation == 0] <- per$id[per$generation == 0] # The easy part
nex <- per$generation == 1
per$tree[nex] <- per$tree[match(per$mid[nex], per$id)] # Kids are of the same 'tree' as mom.
```

Got the idea, the rest by looping over generations 2-10:

```{r looprest}
for (gen in 2:10){
    nex <- per$generation == gen
    per$tree[nex] <- per$tree[match(per$mid[nex], per$id)]
}
head(per[, c("id","mid", "gmid", "generation", "tree")])
```

A summary of what has been done so far:

```{r sumall}
summary(per)
save(per, file = "data/per.rda")
```

And saved for future analysis.








