---
title: "Get indata, second attempt"
author: "Göran Broström"
date: "July 11, 2015"
output: html_document
---

# The time-varying stuff

We start with the *observation* file in the **skel14** package. It has many inconsistencies that need to be fixed.

```{r observation}
library(skel14)
library(dplyr)
per <- filter(person, FODDAT > 18500000 & FODDAT < 18700000)
per <- filter(per, FODDAT %% 100 != 0)
per <- filter(per, (FODDAT %/%100) %% 100 !=0)
dim(per)
obser <- filter(observation, ID %in% per$ID)
obser <- arrange(obser, ID, STARTDAT)
obser <- select(obser, -NOFRS, -ORT, -ORTNMN, -PID, -NARVARO_GEN, -PERS_LOPNR)
length(unique(obser$ID))
id <- unique(obser$ID[obser$STARTDAT < 18500101 | obser$SLUTDAT < 18500101])
length(id)
obser <- filter(obser, !(ID %in% id))
dim(obser)
length(unique(obser$ID))
summary(obser)
id <- unique(obser$ID[obser$STARTDAT > obser$SLUTDAT])
obser <- filter(obser, !(ID %in% id))
length(unique(obser$ID))
```

```{r bplo1}
oj <- obser$STARTDAT[obser$STARTDAT > 16000000] %% 100
barplot(table(oj))
```

Too many days == 0.
Now let us fix invalid dates, ending with "00" (change to "01"), "00xy" (change to "01xy")

```{r fixdates1}
obser$STARTDAT[obser$STARTDAT %% 100 == 0] <- obser$STARTDAT[obser$STARTDAT %% 100 == 0] + 1
sum(is.na(tillTid(obser$STARTDAT)))
obser$STARTDAT[(obser$STARTDAT %/% 100) %% 100 == 0] <- obser$STARTDAT[(obser$STARTDAT %/% 100) %% 100 == 0] + 101
sum(is.na(tillTid(obser$STARTDAT)))
obser$SLUTDAT[obser$SLUTDAT %% 100 == 0] <- obser$SLUTDAT[obser$SLUTDAT %% 100 == 0] + 1
sum(is.na(tillTid(obser$SLUTDAT)))
obser$SLUTDAT[(obser$SLUTDAT %/% 100) %% 100 == 0] <- obser$SLUTDAT[(obser$SLUTDAT %/% 100) %% 100 == 0] + 101
sum(is.na(tillTid(obser$SLUTDAT)))
```

Is this better?

```{r bplo2}
oj <- obser$STARTDAT[obser$STARTDAT > 16000000] %% 100
barplot(table(oj))
```

Oops, the second day in a month is *very* popular, an artefact of our crude elimination of missing dates.

OK, we remove all individuals with any negative interval (already done, btw, but before the manipulation of missing dates):

```{r remneg0}
id <- unique(obser$ID[obser$STARTDAT - obser$SLUTDAT > 0])
obser <- obser[!(obser$ID %in% id), ]
length(unique(obser$ID))
```

Now remove some unnecessary variables:

```{r remunn}
obser <- select(obser, -STARTDAT_MIN, -STARTDAT_MAX, -SLUTDAT_MIN, -SLUTDAT_MAX)
summary(obser)
```

Now disentangle 'TYP == 99'; CIVST or SOCST change?

```{r disent99}
obser <- rc(obser)
obser$ch_civst <- logical(NROW(obser))
obser$next_civst <- c(obser$CIVST[-1], NA)
is.na(obser$next_civst) <- (obser$lopnr == obser$antrec)
obser$ch_socst <- logical(NROW(obser))

obser$next_socst <- c(obser$SOCST[-1], NA)
is.na(obser$next_socst) <- (obser$lopnr == obser$antrec)
obser$ch_civst <- (obser$lopnr < obser$antrec) & (obser$CIVST != obser$next_civst)
obser$ch_socst <- (obser$lopnr < obser$antrec) & (obser$SOCST != obser$next_socst)
with(obser, table(ch_socst, ch_civst, useNA = "ifany"))
obser$move_in <- obser$STARTTYP %in% 4:6
obser$move_out <- obser$SLUTTYP %in% 4:6
obser$prev_move_out <- c(NA, obser$move_out[-NROW(obser)])

is.na(obser$prev_move_out) <- (obser$lopnr == 1)
with(obser[obser$lopnr > 1, ], table(prev_move_out, move_in, useNA = "ifany"))
```

So, there are some individuals with non-consistent migration patterns. We simply remove them

```{r remnoncon}
id <- obser$ID[with(obser, (prev_move_out & !move_in) | (!prev_move_out & move_in))]
obser <- obser[!(obser$ID %in% id), ]
length(unique(obser$ID))
with(obser[obser$lopnr > 1, ], table(prev_move_out, move_in, useNA = "ifany"))
obser$prev_SLUTDAT <- c(NA, obser$SLUTDAT[-NROW(obser)])
is.na(obser$prev_SLUTDAT) <- obser$lopnr == 1
with(obser, summary(STARTDAT - prev_SLUTDAT))
weq <- obser$ID[!is.na(obser$prev_SLUTDAT) & obser$STARTDAT < obser$prev_SLUTDAT]
obser <- obser[!(obser$ID %in% weq), ]
dim(obser)
length(unique(obser$ID))
```

So far, so good. Any inconsistencies left?

```{r incon}
koll <- filter(obser, lopnr >= 2 & !move_in)
sum(koll$STARTDAT != koll$prev_SLUTDAT)
```

# The files

We try to define *events* by *SLUTTYP*. However, the *first entrance* is an event which must be taken into account, similarly every *in-migration* (for people who out-migrate and return).

## The variables.

*Time-fixed* variables (sex, birthdate, parity, etc) are kept in a separate file named **per**.

```{r showper}
per <- per[per$ID %in% obser$ID, ]
dim(per)
```

The *time-varying* variables in our example are *dead*, *SOCST*, *CIVST*, and *NoOfKids*. We also have time-varying presence, guided by in- and out-migration. First we set up the *varsetup* file:

```{r varsetup}
varsetup <- data.frame(name = c("dead", "SOCST", "CIVST", "NoOfKids", "atRisk"), onStart = c(FALSE, rep(TRUE, 4)))
varsetup
```

Note that the proposed fields 'DateFormat' and 'NoFillDown' are redundant: No time-varying variables are dates, and only events (onStart = FALSE) are 'Not Filled Down'.

Next we need a file with *start values* of the time-varying variables. We get it from the first record of each individual:

```{r startval}
starting <-
    obser %>%  
    filter(lopnr == 1) %>%
    select(ID, STARTDAT, SOCST, CIVST)

starting$dead <- FALSE
starting$migration <- 0
starting$childBirth <- 0
starting$atRisk <- TRUE
head(starting)
##save(starting, file = "starting.rda")
```

Next we prepare for the creation of the *episodesFile* by concentrating on *SLUTDAT*. Note though that moving in also is an event with date given by *STARTDAT*, so we need to duplicate some inmigs.

```{r minmig}
obser$dead <- obser$SLUTTYP == 2
minn<- obser[obser$move_in & obser$lopnr > 1, ]
minn$date <- minn$STARTDAT
minn$move_out <- FALSE
##minn$event <- minn$STARTTYP
mout <- obser
mout$date <- mout$SLUTDAT
mout$move_in <- FALSE
##mout$event <- mout$SLUTTYP
out <- rbind(mout, minn)
out <- arrange(out, ID, date)
out <- rc(out) # New 'lopnr' and 'antrec'
head(out)
```

Now, **out** is the file to create the **chronicle** file from. Relevant date is given by the just created variable *date*.
```{r testaout}
event <- with(out, ch_civst | ch_socst | move_in | move_out | dead)
sum(event)
dim(out)
out <- out[event, ]
dim(out)
```

## The chronicle file

First step:

```{r getchron}
chron <- out[, c("ID", "date", "ch_civst", "next_civst", "ch_socst", "next_socst", "move_in", "move_out", "dead")]
oj <- unlist(tapply(chron$dead, chron$ID, cumsum))
chron <- chron[oj <= 1, ]
head(chron)
```

Second step: Run *get_alla*:

```{r get_alla}
source("R/get_chronicle.R")
source("R/get_alla.R")
chronicle <- get_alla(chron)
```

Remains to put on **childBirth** for all guys in *chronicle*:

```{r getbirths}
mothers <- person[person$MID %in% chronicle$id, c("MID", "FODDAT")]
mothers <- mothers[order(mothers$MID, mothers$FODDAT), ]
names(mothers) <- c("id", "Date")
fathers <- person[person$FID %in% chronicle$id, c("FID", "FODDAT")]
fathers <- fathers[order(fathers$FID, fathers$FODDAT), ]
names(fathers) <- c("id", "Date")
births <- rbind(mothers, fathers)
births <- rc(births)
n <- NROW(births)
in_b <- data.frame(id = births$id, 
                   Variable = rep("childBirth"), 
                   Value = births$lopnr, 
                   Date = births$Date,
                   atRisk = 1)
```

Now, join the two data frames and reorder:

```{r joinreorder}
chronicle <- rbind(chronicle, in_b)
chronicle <- arrange(chronicle, id, Date)
head(chronicle)
##save(chronicle, file = "chronicle.rda")
```

And a final cleaning:

```{r cleaning}
id_out <- chronicle$id[chronicle$Date < 18000000]
chronicle <- filter(chronicle, !(id %in% id_out))
names(starting) <- c("id", "Date", "socst", "civst", "dead", "migration", "childBirth", "atRisk")
starting <- starting[starting$id %in% chronicle$id, ]
names(chronicle) <- c("ID_I", "Type", "Value", "Date", "atRisk")
chronicle$Type <- as.character(chronicle$Type)
save(chronicle, file = "chronicle.rda")
indx <- match(starting$id, person$ID)
starting$birthdate <- person$FODDAT[indx]
save(starting, file = "starting.rda")
```