---
title: "IDS, first attempt"
author: "Göran Broström"
date: "27 Jun 2015"
output: pdf_document
---

# Replicating an example

Example data are extracted from the excel file "DemoDatabase.xlsx". It contains several sheets, and we read them one by one with the aid of the **R** package *readxl*. 

```{r setup, include = FALSE}
knitr::opts_chunk$set(comment=NA, fig.width=6, fig.height=6)
options(width = 100)
```

## Reading the Excel file

```{r readx}
library(readxl)
nom <- excel_sheets("data/DemoDatabase.xlsx") # Get the names of the sheets
for (sheet in nom){
   assign(sheet, read_excel("data/DemoDatabase.xlsx", sheet))
}
rm(sheet)
```
We have now got `r length(nom)` data frames with the same names as the sheets they were extracted from. 

```{r ls}
ls()
```

Let us take a look att them. At the same time we will give them more practical names, and also prepare them for easy use with the *dplyr* package (*tbl_df*). First, a note on dates.

## Fix dates

Dates are given by *three* variables, Day, Month, and Year. We transform these to one variable, defined as "years since '1 January the year 1 B.C.'". for instance, 1 January 1801 is 1801.001,
2 July 1879 is 1879.499, etc. Call it "date", "startdate", or "enddate" depending on context.
The function *toYear* does this and also removes the now redundant *Day*, *Month*, *Year* and friends.
This is my "quick fix". 

However, I am wondering why we do not use a standard date-time format, for instance *ISO-8601*. The variable *DayFrac*, appearing in some files, seems to indicate that in some cases information of time-of-day is available. Are the values of *DayFrac* chosen randomly or based on actual knowledge of which came first? 

In R, dates are easily and transparantly treated, for instance

```{r transdat}
x = as.Date("1864-2-29")
y = as.Date("1904-2-29")
c(x, y)
(z = as.numeric(y - x)) # Time difference in days
z / 365.2425 # Time difference in years (1900 was _not_ a leap year)
```


## CON_MET_Import (cmi)

```{r con_met_import, message = FALSE}
library(dplyr)
cmi = tbl_df(CON_MET_Import)
rm(CON_MET_Import)
cmi
```

Of less interest for the time being.

## CONTEXT_CONTEXT (concon)

```{r context_context}
concon <- tbl_df(CONTEXT_CONTEXT)
rm(CONTEXT_CONTEXT)
concon
```


Also less interesting for now.

## CONTEXT_EXT (conext)

From this file we remove the less interesting variables *Id_D*, *DayFrac*, and *Date_type* (besides being less interesting, they are also non-varying).

```{r context_ext, message = FALSE}
conext <- tbl_df(CONTEXT_EXT)
rm(CONTEXT_EXT)
source("R/creDate.R")
conext = creDate(conext)
conext = select(conext, -Id_D, -DayFrac, -Date_type)
conext
```


Also less interesting for now. And all of *startdate* and *enddate* are missing ...

## CONTEXT (con)

```{r context}
con <- tbl_df(CONTEXT)
rm(CONTEXT)
head(con)
```

May be of interest ...

## IND_MET_Import (imi)

We remove the variable *Source*, since it is all missing, and *Explanation*, since it is redundant. *Extract* is also removed (right or wrong?).

```{r ind_met_import}
imi = tbl_df(IND_MET_Import)
rm(IND_MET_Import)
imi = select(imi, -Source, -Explanation, -Extract)
imi
with(imi, table(Type))
```

Probably of less interest for the time being.

## INDIV_CONTEXT (indcon)

```{r indiv_context}
indcon = tbl_df(INDIV_CONTEXT)
rm(INDIV_CONTEXT)
indcon = creDate(indcon)
indcon
```

This is a more interesting (so far) table; it contains individual data in the form of calendar time intervals describing to which context an individual belongs.

## INDIV_INDIV (indind)

```{r indiv_indiv}
indind = tbl_df(INDIV_INDIV)
rm(INDIV_INDIV)
indind
```

No timestamps, so this looks like time-constant relations (but what about a "servant"?).

## INDIVIDUAL (ind)

```{r individual}
ind = tbl_df(INDIVIDUAL)
##rm(INDIVIDUAL)
ind <- creDate(ind)
summary(ind)
```

Look at *startdate* and *enddate*: all missing. Let us remove them:

```{r rmvars}
ind = select(ind, -startdate, -enddate)
ind
```

## INDIVIDUAL_EXT (indext)

We remove some uninteresting variables ... and *startdate* and *enddate*, which are all missing.
```{r individual_ext}
indext <- tbl_df(INDIVIDUAL_EXT)
rm(INDIVIDUAL_EXT)
indext = creDate(indext)
indext <- select(indext, -Id_D, -Source, - startdate, -enddate,
                 -Value_Id_C)
indext
with(indext, table(Id_I, Type, useNA = "ifany"))
```

## Labels (labels)

```{r labels}
labels = tbl_df(Labels)
rm(Labels)
labels
```
Historical reminiscence: distinguishing between labels and values.

## METADATA (meta)

```{r metadata}
meta = tbl_df(METADATA)
rm(METADATA)
head(meta)
```

Need to check what this is ...




# Creating a chronological event file

The two data frames *ind* and *indext* contain the necessary information. We first tidy each separately, then join them.

<!--Seems natural to start with the data frame *ind*. The variables *startdate* and *enddate* are all missing(!), so let us remove them, and also remove the boring variables Id_D and Value_Id_C. We can always change our mind later, if necessary.

```{r timestamps}
ind$startdate <- ind$enddate <- NULL
ind$Id_D <- ind$Value_Id_C <- NULL
```
-->

## The data frame *ind*

We start with the data frame *ind* and list all events in chronological order *within individuals*:

```{r arrange ind}
ind <- arrange(ind, Id_I, date)
ind
with(ind, table(Date_type, Type))
```

This is not a *tidy* data frame (variables in columns, observations in rows). It is practical to distinguish between *time-fixed* (known at birth) and *time-varying* covariates. The latter are best described as *events*, that indices a change in the individual's covariate setup at the time of the event. The variable *Type* seems to indicate a time-fixed covariate (in the given sense), but that is not completely true: *Start_observation* and *End_observation* are clearly time-varying, and can occur more than once. (A note on *time scales*: there is *calendar time* and there is *age*. When we talk about time-fixed we mean on the age scale.)

So, the time-fixed information consists of *birth_date*, *birthplace*, and *sex*. We gather this information in a data frame named *personal*:

```{r creper}
personal <- filter(ind, Type %in% c('Birth_date', 'Birth', 'Sex', 'Birth_location'))
personal
```

It seems as if the distinction between *Birth* and *Birth_date* is that *Birth* is the event that triggers *Start_observation*. Therefore we skip the records with *Type == 'Birth'*. *Birth_location* is all-missing so we let it go as well.

```{r skipbirth}
personal <- filter(personal, !(Type %in% c('Birth', 'Birth_location')))
personal
```

We do not need *Date_type* any longer, and when we give *Birth_date* its proper *Value*, we can remove *date*.

```{r propval}
personal$Value[is.na(personal$Value)] <- round(personal$date[is.na(personal$Value)], 3)
personal <- select(personal, Id_I, Type, Value)
personal
```

This data frame is untidy in a very structured way, which makes it easy to make it tidy: We want a data frame with three variables (columns), *Id_I*, *Birthdate*, and *Sex*. 

```{r fixtidy}
per = data.frame(Id_I = unique(personal$Id_I),
                 Birthdate = round(as.numeric(personal$Value[personal$Type == "Birth_date"]), 3),
                 Sex = personal$Value[personal$Type == "Sex"])
per
```

That's it! now we can "tidy" the data frame *ind* by removing all records referring to *birthdate* or *sex*.

```{r tidyind}
ind <- filter(ind, !(Type %in% c("Birth", "Birth_date", "Birth_location", "Sex")))
ind
with(ind, table(Type, Value, useNA = "ifany"))
```

We want 'Type == "Start_observation"' to be the first record and 'Type == "End_observation"' for each individual. At the same time we get rid of the redundant variables *Date_type* and *Source*.

```{r sten}
ind <- arrange(ind, Id_I, date, Type != "Start_observation")
ind <- arrange(ind, Id_I, date, Type == "End_observation")
ind = select(ind, -Date_type, -Source)
ind
```



## The data frame *indext*

Next data frame to contain dates is *indext*. We sort it similarly and remove the columns *DayFrac*, and *Date_type*:

```{r sortindext}
indext <- arrange(indext, Id_I, date)
indext <- select(indext, -DayFrac, -Date_type)
indext
with(indext, table(Type, Value, useNA = "ifany"))
```

The only interesting Types are *Civil_status* and *ChildBirth* (the other follow or can be reconstructed later), so we remove them.

```{r removesome}
indext <- filter(indext, Type %in% c("Civil_status", "ChildBirth"))
indext
```


## Join the data frames *ind* and *indext*

We call the result *indiv*:

```{r joinin}
ind$Id_I <- as.character(ind$Id_I)
indext$Id_I <- as.character(indext$Id_I)
indext$Value = as.character(indext$Value)
indiv = full_join(ind, indext)
indiv <- arrange(indiv, Id_I, date, Type != "Start_observation")
indiv <- arrange(indiv, Id_I, date, Type == "End_observation")
indiv
```

Now we turn *indiv* into a description of *intervals* (and remove those *starting* with *Death* or *End_observation*):

```{r turniv}
indiv$exitType <- c(indiv$Type[-1], NA)
indiv$exitValue <- c(indiv$Value[-1], NA)
indiv$exitdate <- c(indiv$date[-1], NA)
indiv <- filter(indiv, !(Type %in% c("Death", "End_observation")))
indiv
```

Time to add on *birthdate* and *sex*. We can also safely remove the now redundant variables *date* and *exitdate*:

```{R addonsex}
indx <- match(indiv$Id_I, per$Id_I)
indiv$birthdate <- per$Birthdate[indx]
indiv$sex <- per$Sex[indx]
indiv$enter = round(with(indiv, date - birthdate), 3)
indiv$exit = round(with(indiv, exitdate - birthdate), 3)
indiv = select(indiv, -date, -exitdate)
summary(indiv)
indiv
with(indiv, table(exitType))
```

With so few deaths there is not much of survival analysis to be done, but we can at least plot an estimated survival curve (with 95% confidennce limits):

```{r surv}
library(eha)
ii <- filter(indiv, exit > enter)
with(ii, plot(Surv(enter, exit, exitType == "Death"), fn = "surv"))
```

In a female-fertility study we would select females and use *exitType == ChildBirth* as event.

```{r fertst}
fem <- filter(indiv, sex == "Female", exit > enter)
fit <- coxph(Surv(enter, exit, exitType == "ChildBirth") ~ 1, data = fem)
plot(survfit(fit), fun = "cumhaz")
```

The figure shows the cumulative hazards function of child births (with 95% confidence interval).


## Putting on *civst*

Civil status, *civst* is a time-varying variable that we now can construct. We assume the value *not available* (NA) at start. The same procedure is applied for Occupation (*occ*). After that, *Type* and *Value* are no longer needed.

```{r civston}
indiv$civst = NA
indiv$civst[indiv$Type == "Civil_status"] <- indiv$Value[indiv$Type == "Civil_status"]
indiv$occ <- NA
where <- indiv$Type == "Occupation"
indiv$occ[where] <- indiv$Value[where]
indiv = select(indiv, -Type, -Value)
indiv
```

For *civst* and *occ*, we want to replace 'trailing' NA's by the value preceding them.

```{r trailna}
fill_in <- function(x){
    where <- which(!is.na(x))
    val <- x[where]
    m <- length(where)
    n <- length(x)
    if (m == n) return(x)
    for(i in seq_along(where)){
        x[-(1:(where[i] - 1))] <- val[i]
    }
    x
}

indiv <- indiv %>% group_by(Id_I) %>%
    mutate(occ = fill_in(occ), civst = fill_in(civst))
    
indiv
```

Now some cleaning is what remains: We want labels for *civst*, and also to remove zero-length intervals.

```{r remzero}
indiv$civst[!is.na(indiv$civst) & indiv$civst == 1] <- "unmarried" 
indiv$civst[!is.na(indiv$civst) & indiv$civst == 2] <- "married" 
indiv$civst[!is.na(indiv$civst) & indiv$civst == 3] <- "widow"
indiv <- filter(indiv, exit > enter)
indiv
``` 

We are done. The number of individuals in the final data frame is `r length(unique(indiv$Id_I))`.

# Conclusion

This was fairly easy to accomplish. 
The real challenge is to write a general package for extracting useful data from an IDS data base. 
