---
title: "IDS, fourth attempt"
author: "Göran Broström"
date: "June 19, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(cache=TRUE,  echo=TRUE)
```

# Episodes file creator

This version four of my attempts is essentially equal to version three *(IDS3.Rmd)*, but with new data, the two files created by Luciana's program for the Leuven 2016 conference (infant mortality). The filenames are *ESHD16/Chronicle.dta* and *ESHD16/VarSetup.dta*. 

We work along the lines of Luciana's episodes file creator for Stata, but we take a slightly different approcah in that we first extract the *time-fixed* (called *Invariant* in the *Transition column of the *VarSetup* file) variables from the chronicle file. The *VarSetup* data frame is not needed after that. 

We  *extend the definition of an *event* and consider 

* anything that enforces a change in a time-varying variable to be an event. 

* An event causes a *status change*,

* but there is one *special event*, the final one denoted by *End* in the *Transition* column of the *Varsetup* file. The other events are denoted *Start*

The idea is simply: In order to describe an individual's life history,
we need

* Her **status at start** (birth, first seen, etc)
* **Change** of status: How and when.

## Downloading and loading the necessary files

The input files are in the local directory *./ESHD16*.

```{r download}
library(haven)
chronicle <- read_dta("ESHD16/Chronicle.dta")
varsetup <- read_dta("ESHD16/VarSetup.dta")
```

## The varsetup data frame

The *varsetup* data frame in our example looks like this:

```{r lookatvarsetup}
varsetup
```

The *Type* *Infant_death* is the primary event of interest (denoted *End* in the *Transition* field). The other time-varying variables in our study are *Pregnant* (note to self: a *bad habit* variable), *PreviousInfandAlive*, *YoungerSiblingPresent*, and *At_risk* (denoted *Start* in the *Transition* column). 

Only *Start* and *End* variables will be kept in the *chronicle* data frame, the rest (*Invariant* transition) will be kept in a data frame called *Personal*.

## The chronicle data frame

Let's have a look at the *chronicle* data frames:

```{r thechron}
head(chronicle)
```


If is of interest to look closer at the  **Type** variable:

```{r tabla, echo=FALSE}
(vars <- levels(as.factor(chronicle$Type)))
```

Note the values of the variable **Type**; they will be the the new time-varying covariates, and they must all be described in the *varsetup* data frame. We check that:

```{r checkthat}
notthere <- !(vars %in% varsetup$Type)
if (sum(notthere)){
    cat(paste("Not in VarSetup:", vars[notthere], "\n"))
}else{
    cat("Chronicle Type variable OK\n")
}
```

### Creating the *personal* data frame

Now we can extract the *"Invariant"* covariates from the *chronicle* data frame and store them in the *personal* data frame. We also *sort* it by *Id_I* and *Type*. 

```{r extractpersonal}
toPers <- varsetup$Type[varsetup$Transition == "Invariant"]
personal <- chronicle[chronicle$Type %in% toPers, ]
personal <- personal[with(personal, order(Id_I, Type)), ]
```

We take a look at *personal*


```{r takealookat}
head(personal)
```

we see that *Value* is missing on "dates"; instead there is a *time stamp*. We create a *Value* (date) of the time stamp and store it in *Value*. Then the time stamps are removed, not needed here.

```{r removepersonaltimestamp}
insert <- personal$Value == ""
datumar <- with(personal[insert, ], paste(Year, Month, Day, sep = "-"))
personal$Value[insert] <- datumar
personal$Year <- personal$Month <- NULL
personal$Day <- personal$DayFrac <- NULL
```

Finally, we need to *tidy* the *personal* data frame: A tidy data frame has variables in columns and observations in rows.

```{r tidypersonal}
library(tidyr)
personal <- spread(personal, Type, Value, convert = TRUE)
```

We take a look at it (first six)

```{r lookatpersonal}
head(as.data.frame(personal))
```

### The (rest of the) *chronicle* data file

The rest of the *chronicle*, after *personal* is extracted, is given by

```{r restchronicle}
chronicle <- chronicle[!(chronicle$Type %in% toPers), ]
chronicle <- chronicle[with(chronicle, order(Id_I, Year, Month, Day, Day, DayFrac)), ]
```

where we also put events within individuals in chronological order. It may look like this.

```{r looklikethis}
head(chronicle)
```

Next we want to *tidy* the *chronicle* data frame:

```{r spreadchronicle}
##chronicle$row <- 1:NROW(chronicle)
chronicle$DayFrac[36141] <- 0.02
chronicle <- spread(chronicle, Type, Value, convert = TRUE)
chronicle$datum <- with(chronicle, paste(Year, Month, Day, sep = "-"))
chronicle$datum <- as.Date(chronicle$datum)
##chronicle$Year <- chronicle$Month <- NULL
##chronicle$Day <- NULL
```

```{r newc}
(nv <- unique(chronicle$Type))
chronicle[, nv] <- NA
head(chronicle)
```

The new variables are in place, but they have no values yet ('NA' means 'Not Available', or 'missing').

## Creating the 'episodes' data frame

The episodes data frame is created from the chronicle data frame. We let 

* the **Date** on each row correspond to the **end** of a corresponding interval. 

* The **start** date will be taken from the **previous record** of the individual. 

* For the **first** record we pick the start date from the **starting** data frame.

First we ensure that the two data frames are sorted in the same way, after
**id**. The chronicle file is to be sorted after date within individual. We
also change the variable name **ID_I** in chronicle to **id*; simpler and
the same name as in the **starting** data frame.

```{r manipulate}
names(chronicle)[1] <- "id"
library(dplyr)
chronicle <- arrange(chronicle, id, Date)
starting <- arrange(starting, id) # Only one record per id.
```

Then we introduce two help variables, **lopnr** and **antrec**. They will
help us keeping track of where in an individual's records we presently are,
especially if we are at the first or last record. The function **rc** from
the package **skum* help us with that. 

```{r recnum}
source("R/rc.R")
chronicle <- rc(chronicle)
head(chronicle)
```

Next we create the important variables 

* **enter** and **exit**. They are the 
* *age* of the individual at start and end of an interval.  

First we put on the birth date of each individual on each of her or his interval. Then we use that together with the function **tillTid** to create the age (in years) at exit.

```{r birthdate}
source("R/tillTid.R")
indx <- match(chronicle$id, starting$id)
chronicle$birthdate <- starting$birthdate[indx]
chronicle$exit <- tillTid(chronicle$Date) - tillTid(chronicle$birthdate)
chronicle$enter <- c(NA, chronicle$exit[-NROW(chronicle)])
chronicle$enter[chronicle$lopnr == 1] <- tillTid(starting$Date) - tillTid(starting$birthdate) # dangerous!
head(chronicle)
```

The next step is to put on start values. Note that they are valid until an event changes things, sot they are put in on all records for each individual.

```{r startv}
stvar <- nv[nv %in% names(starting)]
chronicle[, stvar] <- starting[indx, stvar]
episodes <- chronicle
head(episodes)
```

With start values in place, we roll down the time ladder and add the rest. It takes a fairly long time (6--7 minutes) due to non-optimized code, so the first time this code is run, the created data file is saved on the hard disk, and the next time it is simply read from disk, if present.

```{r goon}
if (file.exists("episodes.rda")){
    load("episodes.rda")
}else{
    source("R/fill.R")
    episodes <- fill(episodes)
    save(episodes, file = "episodes.rda")
}
```

Now the **episodes** data frame looks like this:

```{r lookalike}
head(episodes)
summary(episodes)
```

From the summary it is obvious that some errors in the raw data are present: 

* Intervals start before birth, 
* Missing values in dates. 

We simply remove all individuals with any of these defects (for this demonstration.

```{r removal}
weq <- unique(episodes$id[is.na(episodes$enter) | episodes$enter < 0])
episodes <- episodes[!(episodes$id %in% weq), ] 
summary(episodes)
```

So we removed `r length(weq)` individuals (with all their records). Still some trouble, one record with missing **exit**:

```{r missex}
id <- episodes$id[is.na(episodes$exit)]
episodes[episodes$id %in% id, ]
```
obviously a bad date (18910500). Let's make a 'quick fix': 

* Remove all (i.e., just one) records with missing **exit** 

```{r quick}
episodes = episodes[!is.na(episodes$exit), ]
```

Are there any non-positive-length intervals?

```{r nonpos}
(nonpos <- sum(episodes$enter >= episodes$exit))
```

* There are `r nonpos`. We remove them.

```{r remthem}
episodes <- episodes[episodes$exit > episodes$enter, ]
dim(episodes)
length(unique(episodes$id))
```
Still some data left. 

# The time-fixed covariates

Finally, we put on the fixed covariates from **personal**. Note that the first two columns of the file (always) contain **id** and **birthdate**, in that order, so

```{r fixedcov}
indx <- match(episodes$id, personal$id)
vars <- names(personal)[-(1:2)]
episodes[, vars] <- personal[indx, vars]
head(episodes)
```

Some cleaning is appropriate: Keep only records "atRisk", remove some redundant, irrelevant covariates.

```{r remred}
#episodes <- episodes[episodes$atRisk == 1, ]
episodes$Date <- episodes$dead <- episodes$migration <- NULL
episodes$lopnr <- episodes$antrec <- NULL

```

Some "value labels" will be nice:

```{r vallab}
episodes$Type <- as.factor(episodes$Type)
episodes$sex <- factor(episodes$sex, labels = c("man", "woman"))
summary(episodes)
```


# Mortality

Time to perform some analyses, starting with mortality!

```{r morta, cache=TRUE}
episodes$exit <- as.numeric(episodes$exit)
fit <- coxreg(Surv(enter, exit, Type == "dead") ~ sex, data = episodes)
(dr <- drop1(fit, test = "Chisq"))
fit
```

And the plot.

```{r deadplot, echo = FALSE}
fit <- coxph(Surv(enter, exit, Type == "dead") ~ strata(sex), data = episodes)
plot(survfit(fit), mark.time = FALSE, col = c("blue", "red"), xlab = "Age", 
     ylab = "Surviving fraction")
text(25, 0.5, "Women", col = "red")
text(55, 0.8, "Men", col = "blue")
abline(h = 0)
```

# Fertility

```{r fertility}
fit <- coxph(Surv(enter, exit, Type == "childBirth") ~ strata(sex), data = episodes)
plot(survfit(fit), mark.time = FALSE, fun = "cumhaz", col = c("blue", "red"),
     xlab = "Age", ylab = "No of births", xlim = c(16, 60))
text(25, 3, "Women", col = "red")
text(45, 3, "Men", col = "blue")
abline(h = 0)
```

## First birth, women

```{r  firstbi, cache = TRUE}
fit <- coxph(Surv(enter, exit, Type == "childBirth") ~ civst + birthdate, 
             data = episodes[episodes$sex == "woman" & episodes$childBirth == 1, ])
summary(fit)
plot(survfit(fit), mark.time = FALSE, xlab = "Age", xlim = c(17, 40), col = "blue")
abline(h = 0)
```

## First birth interval, women

```{r firstiv, cache = TRUE}
first <- episodes[episodes$sex == "woman" & 
                     episodes$childBirth == 1 &
                     episodes$civst %in% 1:3, ]
first$civst <- factor(first$civst, labels = c("unmarried", "married", "widow"))

indx <- with(first, tapply(id, id))
minEnter <- with(first, tapply(enter, id, min))
minEnter <- minEnter[indx]
first$enter <- first$enter - minEnter
first$exit <- first$exit - minEnter
head(first[, c("id", "enter", "exit", "Type")])
```

Now we get

```{r, firstivsk, cache = TRUE}
fit <- coxph(Surv(enter, exit, Type == "childBirth") ~ strata(civst), data = first)
plot(survfit(fit), fun = "cumhaz", col = c("blue", "red", "black"), 
    mark.time = FALSE, xlim = c(0, 10), xlab = "Years", ylab = "Cumulative hazards", 
    main = "First birth interval")
text(2, 2.5, "married", col = "red")
text(5, 1.0, "widowed", col = "black")
text(7, 0.5, "unmarried", col = "blue")
abline(h = 0)
```


