---
title: "To IDS format"
author: "Göran Broström"
date: "May 6, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(skum)
sobs <- obs[obs$region == "ske", ]
sper <- per[per$region == "ske" & !is.na(per$foddat), ]
sper$frsbobdat[is.na(sper$frsbobdat)] <- sper$foddat[is.na(sper$frsbobdat)]
sper$frsbosdat[is.na(sper$frsbosdat)] <- sper$doddat[is.na(sper$frsbosdat)]
sper <- sper[!is.na(sper$frsbosdat), ]
sper$frsbosdat[sper$frsbosdat < sper$foddat] <- sper$foddat[sper$frsbosdat < sper$foddat]
sper$mid[!is.na(sper$mid) & !(sper$mid %in% sper$id)] <- NA
library(xtable)
library(knitr)
options(stringsAsFactors = FALSE)
```

## The Skellefteå data

In the **R** package *skum*, Skellefteå inhabitants born between 1815 and 1950 are included. Data are given in two sources, a *person* file containing *static* information, i.e., information that is *not* changing over time, and an *observation* file, representing *presence* intervals in the parish, where each time interval contain information fixed in that interval, but information may vary between intervals for the same individual.

### The *person* file

Each individual is represented by one row in the data frame:

```{r oneper}
kable(sper[3:4, c("id", "kon", "mid", "fid", "foddat", "doddat")], row.names = FALSE)
```

There are `r NCOL(per)` variables (columns) in total in *per*, here some of them are selected: *ID, mother's ID, father's ID, birthdate, death date*. No. 9 has full information (on the selected variables), while No. 17 has no links to the parents and his death date is missing, probably due to the fact that he out-migrated and was lost to follow-up.

### The *observation* file

```{r oneobs}
kable(sobs[sobs$id %in% c(9, 17), c("id", "startdat", "starttyp", "slutdat", "sluttyp", "socBranch", "civst", "pid")], row.names = FALSE)
```

No. 9 has many time intervals, due to the fact that he changes civil and social status several times.

## Creation of the INDIVIDUAL table

To save some efforts for this particular event, we concentrate on *mothers* and *children*. First, mothers are all those women who are linked to by the variable *mid* in the *per* file:

```{r mothers}
mothers <- unique(sper$mid[!is.na(per$mid)])
length(mothers)                  
```

Children are those individuals with *mid* in the set *mothers*:

```{r children}
children <- sper$id[(!is.na(sper$mid)) & sper$mid %in% mothers] 
length(children)
```

Now, the union of "mothers" and "children" should make up the content of "INDIVIDUAL":

```{r content}
content <- unique(c(mothers, children))
length(content)
```

According to Luciana's "Program for studying intergenerational ...", each individual in INDIVIDUAL
should be described in six records, representing the *Type*s *Birth_date*, *Death_date*, *Sex*, *Start_observation*, *End_observation*, and *Marriage_date*. We create one data frame for each Type to start with, and then we join them into one data frame INDIVIDUAL.

```{r onebyone}	    
pe <- sper[sper$id %in% content, ]
ind1 <- data.frame(id_I = pe$id,
                   Type = "Birth_date",
                   Value = as.character(pe$foddat))
		   	 
ind2 <- data.frame(id_I = pe$id,
                   Type = "Death_date",
                   Value = as.character(pe$doddat))
		   	 
ind3 <- data.frame(id_I = pe$id,
                   Type = "Sex",
                   Value = pe$kon)
		   	 
ind4 <- data.frame(id_I = pe$id,
                   Type = "Start_observation",
                   Value = as.character(pe$frsbobdat))
		   	 
ind5 <- data.frame(id_I = pe$id,
                   Type = "End_observation",
                   Value = as.character(pe$frsbosdat))
		   	 
ind6 <- data.frame(id_I = pe$id,
                   Type = "Marriage_date",
                   Value = NA)
```

We now have the small problem of getting the *marriage date*. However,
the *first* marriage date is easy to find, it is the minimum of *startdat* for
intervals with *civst == married* (for each individual). The only technical
complication is that not all get married and that some marries more than once.
We play it simple and are satisfied with the first marriage date (if any).

We single out the subset of *sobs* with only married:

```{r mardat}
mar <- sobs[sobs$civst == "married", c("id", "startdat")]
mar <- mar[order(mar$id, mar$startdat), ] # To make sure ...
mar <- mar[!duplicated(mar$id), ]
kable(mar[1:5, ], row.names = FALSE)
```

Remains to put *startdat* in the *Value* column of *ind6* by matching on *id* and *id_I*:

```{r matchon}
ind6$Value <- as.character(mar$startdat[match(ind6$id_I, mar$id)])
kable(ind6[1:5, ], row.names = FALSE)
```

A missing value means no marriage (in our registers).

The final *INDIVIDUAL* is created by joining the six data frames into one and reorder:

```{r INDIVIDUAL}
INDIVIDUAL <- rbind(ind1, ind2, ind3, ind4, ind5, ind6)
INDIVIDUAL <- INDIVIDUAL[order(INDIVIDUAL$id_I), ]
kable(INDIVIDUAL[1:12, ], row.names = FALSE)
```

Remains to create the table INDIV_INDIV:

```{r}
ii <- sper[sper$id %in% children, c("mid", "id")]
ii <- ii[order(ii$mid, ii$id), ]
INDIV_INDIV <- data.frame(id_1 = ii$mid,
                          id_2 = ii$id,
                          Relation = "Mother")
kable(INDIV_INDIV[1:5, ], row.names = FALSE)
```

Remains to save the two created data frames:

```{r saveall}
save(INDIVIDUAL, file = "data/INDIVIDUAL.rda")
save(INDIV_INDIV, file = "data/INDIV_INDIV.rda")
```





